---
title: "Class 8: Halloween Mini-Project"
author: "Jason Hsiao (PID: A15871650)"
format: pdf
---

Importing the Candy Dataset
```{r}
candy <- read.csv("candy-data.csv", row.names=1)
head(candy)

```
What's in the dataset?
```{r}
nrow(candy) #rows in the dataset (candy types)

```
> Q1: How many different candy types are in this dataset?

There are 85 different candy types in the dataset


> Q2: How many fruity candy types are in the dataset?

```{r}
sum(candy$fruity)

```

There are 38 fruity candy types in the dataset.

> Q3: What is your favorite candy in the dataset and what is its `winpercent` value?

```{r}
candy["Peanut M&Ms", ]$winpercent

```
Win % of Peanut M&Ms is 69.5%


> Q4: What is the winpercent value for “Kit Kat”?

```{r}
candy["Kit Kat", ]$winpercent

```
Win % of Kit Kat is 76.8%.


> Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
candy["Tootsie Roll Snack Bars", ]$winpercent

```
Win % of Tootsie Roll Snack Bars is 49.7%.

Using the `skim()` function to get an overview of a given dataset...
```{r}
library("skimr")
skim(candy)

```

> Q6: Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

winpercent values are in whole percentages, while others are in proportions.


> Q7: What do you think a zero and one represent for the candy$chocolate column?

Whether or not the given candy from a given row is (one) and isn't (zero) under the chocolate category.

> Q8: Plot a histogram of winpercent values

```{r}
library(ggplot2)
ggplot(candy, aes(x = winpercent)) +
  geom_histogram()

```

> Q9: Is the distribution of winpercent values symmetrical?

No, it appears to have a right-skew.


> Q10: Is the center of the distribution above or below 50%?

Below 50%.


> Q11: On average is chocolate candy higher or lower ranked than fruit candy?

```{r}
#chocolate candy winpercent
mean(candy$winpercent[as.logical(candy$chocolate)])

#fruit candy winpercent
mean(candy$winpercent[as.logical(candy$fruity)])

```
Chocolate candy is ranked higher than fruit candy!


> Q12: Is this difference statistically significant?

```{r}
#assign 'choc' object as vector of win percentages of chocolate candies
choc <- candy$winpercent[as.logical(candy$chocolate)]
#assign 'fruit' object as vector of win percentages of fruity candies
fruity <- candy$winpercent[as.logical(candy$fruity)]

t.test(choc, fruity)

```
The difference is statistically significant!


> Q13: What are the five least liked candy types in this set?

```{r}
library(dplyr)

candy %>% 
  arrange(winpercent) %>%
  head(5)

```
Nik L Nip, Boston Baked Beans, Chiclets, Super Bubble, Jawbusters are the 5 least liked candies!


> Q14: What are the top 5 all time favorite candy types out of this set?

```{r}
candy %>%
  arrange(-winpercent) %>%
  head(5)
  
```
Reese's Peanut Butter cup, Reese's Miniatures, Twix, Kit Kat, Snickers are the top 5 favorites!


> Q15: Make a first barplot of candy ranking based on winpercent values.

```{r}
ggplot(candy, aes(winpercent, rownames(candy))) +
  geom_col()

```
Reordering by winpercent:
```{r}
#setting color parameters
my_cols = rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"

ggplot(candy, aes(winpercent, reorder(rownames(candy),winpercent))) +
  geom_col(fill = my_cols)

```

> Q17: What is the worst ranked chocolate candy?

Nik L Nip


> Q18: What is the best ranked fruity candy?

Starburst


## Exploring pricepercent...
```{r}
#library(ggrepel)

# How about a plot of price vs win
ggplot(candy) +
  aes(winpercent, pricepercent, label = rownames(candy)) +
  geom_point(col = my_cols)  
  #geom_text_repel(col = my_cols, size = 3.3, max.overlaps = 5)

```

> Q19: Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

```{r}
#ordering pricepercent to see most/least expensive candy
ord <- order(candy$pricepercent, decreasing = FALSE)
head( candy[ord, c(11,12)], n = 5)

```

Tootsie Roll Midgies


> Q20: What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head( candy[ord, c(11,12)], n = 5)

```

Nik L Nip is the most expensive and least popular of the dataset. 


## Correlation Plot
```{r}
library(corrplot)

cij <- cor(candy)
corrplot(cij)

```
> Q21: Make a barplot again with geom_col() this time using pricepercent and then improve this step by step, first ordering the x-axis by value and finally making a so called “dot chat” or “lollipop” chart by swapping geom_col() for geom_point() + geom_segment().

```{r}
#barplot of pricepercent
ggplot(candy) +
  aes(pricepercent, reorder(rownames(candy), pricepercent)) +
  geom_col()

#making it prettier: make a lollipop chart of pricepercent
ggplot(candy) +
  aes(pricepercent, reorder(rownames(candy), pricepercent)) +
  geom_segment(aes(yend = reorder(rownames(candy), pricepercent), 
                   xend = 0), col="gray40") +
    geom_point()

```


> Q22: Examining this plot what two variables are anti-correlated (i.e. have minus values)?

chocolate and fruity


> Q23: Similarly, what two variables are most positively correlated?

chocolate and winpercent


## PCA
```{r}
#setting up PCA object
pca <- prcomp(candy, scale. = TRUE)
summary(pca)

plot(pca$x[, 1:2], col = my_cols, pch = 16)

# Make a new data-frame with our PCA results and candy data
my_data <- cbind(candy, pca$x[,1:3])

p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=my_cols)

p

#putting labels on each of the dots; making it prettier 
#library(ggrepel)

#p + geom_text_repel(size=3.3, col=my_cols, max.overlaps = 7)  + 
  #theme(legend.position = "none") +
  #labs(title="Halloween Candy PCA Space",
       #subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), #other (black)",
       #caption="Data from 538")

#passing plot 'p' to plotly to get interactive plot (removes overlap)
#library(plotly)
#ggplotly(p)

```

## PC Loadings
```{r}
par(mar=c(8,4,2,2))
barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution")

```
> Q24: What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?

fruity, hard, pluribus - yes, it makes sense (they normally come in this form)